Epilepsy GLMM example

### 2022-12-12 src/epil/epil.Rmd
Add data plot, MCMC diagnostics, more R-INLA strategies, and timings

---
title: "Epilepsy example"
author:
- name: Adam Howes
output:
  bookdown::html_document2:
    toc: yes
    toc_float: yes
    toc_collapsed: yes
    df_print: paged
    code_folding: show
    theme: lumen
bibliography: citations.bib
abstract: |
  **Background** @rue2009approximate use the epilespy example from the OpenBUGs manual as demonstration of the integrated nested Laplace approximation method. The model is a Poisson GLMM including both fixed and random effects.

  **Task** We implement the epilepsy example in Stan, `R-INLA`, `TMB`, `glmmTMB`, `tmbstan`, `aghq`, and a custom Laplace marginal version of `aghq`. We then compare the results of the inference methods.
---

@@

@@

```{r}
fit1 <- stan(
  file = "epil.stan",
  data = dat,
  chains = 4,
  warmup = 100,
  iter = 1000,
  control = list(adapt_delta = 0.95),
  refresh = 0
)
```

@@

fit2 <- epil_inla(strat = "gaussian", int_strat = "eb")
fit3 <- epil_inla(strat = "simplified.laplace", int_strat = "eb")
fit4 <- epil_inla(strat = "laplace", int_strat = "eb")

fit5 <- epil_inla(strat = "gaussian", int_strat = "grid")
fit6 <- epil_inla(strat = "simplified.laplace", int_strat = "grid")
fit7 <- epil_inla(strat = "laplace", int_strat = "grid")
```

@@

@@

```{r}
obj <- MakeADFun(
  data = dat,
  parameters = param,
  random = c("beta", "epsilon", "nu"),
  DLL = "epil",
  silent = TRUE
)
```

The objective function `obj$fn` and its gradient `obj$gn` are a function of only the parameters, which in this instance are the logarithms of $\tau_\epsilon$ and $\tau_\nu$.
This can be checked with `names(obj$par)`.
Optimise `obj` using 1000 iterations of the the `nlminb` optimiser, passing in the starting values `start`, objective function `objective` and its derivative `gradient`:

@@

sd_out <- sdreport(
  obj,
  par.fixed = opt$par,
  getJointPrecision = TRUE
)
```

@@

For completeness, we check here that the objective functions as implemented in `TMB` and Stan are identical.
To obtain the `TMB` negative log-likelihood we call `MakeADFun` as before, but now do not specify any parameters to be integrated out.
For Stan, we create an empty model, then use `rstan::log_prob`:

@@

Now test the NLL of the initialisation parameters `param`:

@@

This look different!
Is this a problem?
We can evaluate the NLL at some of the other parameters explored during the MCMC:

@@

```{r}
formula6 <- y ~ 1 + CTrt + ClBase4 + CV4 + ClAge + CBT + (1 | rand) + (1 | Ind)
fit6 <- glmmTMB(formula6, data = Epil, family = poisson(link = "log"))
```

@@

```{r}
fit7 <- tmbstan(obj = obj, chains = 4, refresh = 0)
```

@@

[`aghq`](https://arxiv.org/pdf/2101.04468.pdf) is an R package written by Alex Stringer which implements adaptive Gauss-Hermite quadrature.
It also works smoothly with `TMB` templates.
One (odd!) approach would be to use `glmmTMB` to get the `TMB` template which can be passed the `aghq`.

```{r}
glmm_model_info <- glmmTMB(formula6, data = Epil, family = poisson(link = "log"), doFit = FALSE)

@@

Another, more standard, approach is to use the `TMB` template, `obj`, that we already have:

```{r results='hide'}
fit11 <- aghq::marginal_laplace_tmb(
  obj,
  k = 3,
  startingvalue = c(param$l_tau_epsilon, param$l_tau_nu)
)

fit11_samples <- aghq::sample_marginal(fit8, M = 1000)$samps %>%
  t() %>%
  as.data.frame() %>%
  inf.utils::replace_duplicate_colnames()
```

@@

### Using multiple C++ template

@@

### Using one C++ template

@@

Eventually create function of the form:

@@

@@

```{r}
stan1 <- as.vector(t(summary(fit1)$summary[1:6, c(1, 3)]))
inla2 <- as.vector(t(fit2$summary.fixed[1:6, 1:2]))
inla3 <- as.vector(t(fit3$summary.fixed[1:6, 1:2]))
inla4 <- as.vector(t(fit4$summary.fixed[1:6, 1:2]))
inla5 <- as.vector(t(fit2$summary.fixed[1:6, 1:2]))
inla6 <- as.vector(t(fit3$summary.fixed[1:6, 1:2]))
inla7 <- as.vector(t(fit4$summary.fixed[1:6, 1:2]))
tmb8 <- as.vector(t(data.frame(sd_out$par.random[1:6], sqrt(sd_out$diag.cov.random[1:6]))))
glmmtmb9 <- as.vector(t(summary(fit6)$coefficients$cond[, c("Estimate", "Std. Error")]))
tmbstan10 <- as.vector(t(summary(fit7)$summary[1:6, c(1, 3)]))

fit11_beta_samples <- fit8_samples[, 1:6]
aghq11 <- as.vector(t(data.frame(mean = apply(fit11_beta_samples, 2, mean), sd = apply(fit8_beta_samples, 2, sd))))

@@

beta_i <- function(i) { c(paste0("beta_", i), paste0("sd(beta_", i, ")")) }
rownames(df) <- c(sapply(0:5, beta_i))
colnames(df) <- c("Stan", "EB-G", "EB-SL", "EB-L", "Grid-G", "Grid-SL", "Grid-L", "TMB", "glmmTMB", "tmbstan", "aghq", "adam", "type")

@@

>>>>

---
title: "Epilepsy GLMM example"
author:
- name: Adam Howes
output:
  bookdown::html_document2:
    toc: yes
    toc_float: yes
    toc_collapsed: yes
    df_print: paged
    code_folding: show
    theme: lumen
bibliography: citations.bib
abstract: |
  **Background** @rue2009approximate use the epilespy example from the OpenBUGs manual as demonstration of the integrated nested Laplace approximation method. The model is a Poisson GLMM including both fixed and random effects.

  **Task** We implement the epilepsy example in Stan, `R-INLA`, `TMB`, `glmmTMB`, `tmbstan`, `aghq`, and a custom Laplace marginal version of `aghq`. We then compare the results of the inference methods. This serves to compare inference methods, and check new methods are behaving as expected.
---

@@

```{r fig.cap="It looks like treatment is associated to fewer seizures on average, but of course we haven't accounted for the different covariates, hence the use of a statistical model!"}
Epil %>% ggplot(aes(x =  as.factor(Trt), y = y)) +
  geom_boxplot(alpha = 0.5, outlier = FALSE) +
  geom_jitter(width = 0.1, alpha = 0.5) +
  theme_minimal() +
  labs(x = "Treatment", y = "Number of seizures")
```

@@

```{r}
start_time_fit1 <- Sys.time()

fit1 <- stan(
  file = "epil.stan",
  data = dat,
  chains = 4,
  warmup = 100,
  iter = 1000,
  control = list(adapt_delta = 0.95),
  refresh = 0
)

end_time_fit1 <- Sys.time()
time_fit1 <- end_time_fit1 - start_time_fit1
```

It took `r time_fit1` seconds to fit this model.
All of the $\hat R$ values are below 1.05, so it looks like the MCMC has reached convergence:

```{r}
monitor1 <- rstan::monitor(fit1, print = FALSE)

monitor1 %>%
  as.data.frame() %>%
  tibble::rownames_to_column(var = "parameter") %>%
  ggplot(aes(x = parameter, y = Rhat)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 1.05, linetype = "dashed") +
  labs(x = "Parameter")  +
  theme_minimal() +
  theme(
    axis.title.x = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    legend.position = "bottom"
  )
```

@@

start_time_fit2 <- Sys.time()
fit2 <- epil_inla(strat = "gaussian", int_strat = "eb")
end_time_fit2 <- Sys.time()
(time_fit2 <- end_time_fit2 - start_time_fit2)

start_time_fit3 <- Sys.time()
fit3 <- epil_inla(strat = "simplified.laplace", int_strat = "eb")
end_time_fit3 <- Sys.time()
(time_fit3 <- end_time_fit3 - start_time_fit3)

start_time_fit4 <- Sys.time()
fit4 <- epil_inla(strat = "laplace", int_strat = "eb")
end_time_fit4 <- Sys.time()
(time_fit4 <- end_time_fit4 - start_time_fit4)

start_time_fit5 <- Sys.time()
fit5 <- epil_inla(strat = "gaussian", int_strat = "grid")
end_time_fit5 <- Sys.time()
(time_fit5 <- end_time_fit5 - start_time_fit5)

start_time_fit6 <- Sys.time()
fit6 <- epil_inla(strat = "simplified.laplace", int_strat = "grid")
end_time_fit6 <- Sys.time()
(time_fit6 <- end_time_fit6 - start_time_fit6)

start_time_fit7 <- Sys.time()
fit7 <- epil_inla(strat = "laplace", int_strat = "grid")
end_time_fit7 <- Sys.time()
(time_fit7 <- end_time_fit7 - start_time_fit7)
```

@@

Template Model Builder, or `TMB`, is an R package which implements Laplace approximation using automatic differentiation.
The user writes a C++ file, here `epil.cpp`, which contains the negative log-posterior for the model in question:

@@

```{r}
#' Perhaps you could argue about when the "start" and "end" of the model fitting are
#' For example, is compilation of the TMB template part of the model fitting?
start_time_fit8 <- Sys.time()

obj <- MakeADFun(
  data = dat,
  parameters = param,
  random = c("beta", "epsilon", "nu"),
  DLL = "epil",
  silent = TRUE
)
```

The objective function `obj$fn` and its gradient `obj$gn` are a function of only the parameters, which in this instance are the logarithms of $\tau_\epsilon$ and $\tau_\nu$.
This can be checked with `names(obj$par)`.
We optimise `obj` using 1000 iterations of the the `nlminb` optimiser, passing in the starting values `start`, objective function `objective` and its derivative `gradient`:

@@

sd_out <- sdreport(
  obj,
  par.fixed = opt$par,
  getJointPrecision = TRUE
)

end_time_fit8 <- Sys.time()
(time_fit8 <- end_time_fit8 - start_time_fit8)
```

@@

For completeness, we check here that the objective functions as implemented in `TMB` and Stan are identical.
To obtain the `TMB` negative log-posterior we call `MakeADFun` as before, but now do not specify any parameters to be integrated out.
For Stan, we create an empty model, then use `rstan::log_prob`:

@@

For example, we can test the NLL of the initialisation parameters `param`:

@@

These look different!
Is this a problem?
We can evaluate the NLL at some of the other parameters explored during the MCMC:

@@

```{r}
start_time_fit9 <- Sys.time()
formula9 <- y ~ 1 + CTrt + ClBase4 + CV4 + ClAge + CBT + (1 | rand) + (1 | Ind)
fit9 <- glmmTMB(formula9, data = Epil, family = poisson(link = "log"))
end_time_fit9 <- Sys.time()
(time_fit9 <- end_time_fit9 - start_time_fit9)
```

@@

```{r} 
start_time_fit10 <- Sys.time()
fit10 <- tmbstan(obj = obj, chains = 4, refresh = 0)
end_time_fit10 <- Sys.time()
(time_fit10 <- end_time_fit10 - start_time_fit10)
```

@@

[`aghq`](https://arxiv.org/pdf/2101.04468.pdf) is an R package written by Alex Stringer which implements adaptive Gauss-Hermite quadrature.
It also works smoothly with `TMB` templates.

One (non-standard!) approach would be to use `glmmTMB` to get the `TMB` template which can then be passed to `aghq`.

```{r}
glmm_model_info <- glmmTMB(formula9, data = Epil, family = poisson(link = "log"), doFit = FALSE)

@@

This would be useful if we didn't have a `TMB` template written, but we do, so let's just use that instead!

```{r results='hide'}
start_time_fit11 <- Sys.time()
fit11 <- aghq::marginal_laplace_tmb(
  obj,
  k = 3,
  startingvalue = c(param$l_tau_epsilon, param$l_tau_nu)
)

fit11_samples <- aghq::sample_marginal(fit11, M = 1000)$samps %>%
  t() %>%
  as.data.frame() %>%
  inf.utils::replace_duplicate_colnames()
end_time_fit11 <- Sys.time()
(time_fit11 <- end_time_fit11 - start_time_fit11)
```

@@

### Empirical Bayes with multiple C++ template

@@

### Empirical Bayes with one C++ template

@@

Create function which takes as input the index `i` (from `index_controller`), the results of optimisation for $\theta$, and a grid of values for $x_i$ to evaluate at:

@@

### AGHQ with one C++ template

```{r}
i <- 1
dat$toggle <- index_controller[i, "toggle"]
dat$i <- index_controller[i, "i"]
```

@@

```{r}
stan1 <- as.vector(t(summary(fit1)$summary[1:6, c(1, 3)]))
inla2 <- as.vector(t(fit2$summary.fixed[1:6, 1:2]))
inla3 <- as.vector(t(fit3$summary.fixed[1:6, 1:2]))
inla4 <- as.vector(t(fit4$summary.fixed[1:6, 1:2]))
inla5 <- as.vector(t(fit5$summary.fixed[1:6, 1:2]))
inla6 <- as.vector(t(fit6$summary.fixed[1:6, 1:2]))
inla7 <- as.vector(t(fit7$summary.fixed[1:6, 1:2]))
tmb8 <- as.vector(t(data.frame(sd_out$par.random[1:6], sqrt(sd_out$diag.cov.random[1:6]))))
glmmtmb9 <- as.vector(t(summary(fit9)$coefficients$cond[, c("Estimate", "Std. Error")]))
tmbstan10 <- as.vector(t(summary(fit10)$summary[1:6, c(1, 3)]))

fit11_beta_samples <- fit11_samples[, 1:6]
aghq11 <- as.vector(t(data.frame(mean = apply(fit11_beta_samples, 2, mean), sd = apply(fit11_beta_samples, 2, sd))))

@@

beta_i <- function(i) { c(paste0("beta_", i), paste0("sd(beta_", i, ")")) }
rownames(df) <- c(sapply(0:5, beta_i))
method_names <- c("Stan", "EB-G", "EB-SL", "EB-L", "Grid-G", "Grid-SL", "Grid-L", "TMB", "glmmTMB", "tmbstan", "aghq", "adam")
colnames(df) <- c(method_names, "type")

@@

```{r}
time_df <- data.frame(
  time = c(time_fit1, time_fit2, time_fit3, time_fit4, time_fit5, time_fit6, time_fit7, time_fit8, time_fit9, time_fit10, time_fit11),
  method = head(method_names, n = 11)
)

time_df %>% ggplot(aes(x = method, y = time)) +
  geom_col() +
  theme_minimal() +
  labs(x = "Method", y = "Time taken")
```
